Relative parabolic factorizations of Chevalley groups

### 2015-08-23 paper.tex
Minor changes

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative versions of the following three kinds of decompositions:
\begin{itemize}
\item \emph{Gauss decomposition,} i.\,e. a decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item \emph{Bass---Kolster} decomposition $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
\item \emph{Dennis---Vaserstein} decomposition $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
      where $\EP_i$ denotes the subgroup generated by root unipotents $t_\alpha(\xi)$ contained in $\Par_i$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

On the other hand, much less is known about the width of relative groups $\G(\Phi, R, I)$. 
It is a classical result of J.~Tits and L.~Vaserstein that the relative elementary group $\E(\Phi, R, I)$ is generated by elements $z_\alpha(s,\xi)=t_{\alpha}(s)^{t_{-\alpha}(\xi)}$, see Theorem~\ref{theorem:Tits-Vaserstein} below.
Denote by $\mathcal{Z}$ the set of all $z_\alpha(s, \xi)$, $s\in I$, $\xi \in R$. 
Our first application of parabolic factorizations is the following estimate of the width of $\G(\Phi, R, I)$ in terms of $\mathcal{Z}$.

@@

The second application concerns subsystem factorizations of Chevalley groups formulated in terms of submaximal subgroups of type $\rA_\ell$.
In case when $R=I$ is a field this question was studied by N.~Nikolov in \cite{NikProdDecomp}. 
\begin{thm*} Let $G$ be a classical Chevalley group (of normal or twisted type) of rank $n$ over a finite field.
Then $G$ equals the product of at most $200$ conjugates of $\SL_n$. \end{thm*}

@@

Another applications of parabolic factorizations developed in this paper are subsystem factorizations in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{NikProdDecomp}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.
In~\cite{VavKovSL2} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, Gauss decomposition (see~\cite{Sm12}) implies a similar factorization for the elementary subgroup $\E(\Phi, R)$ with $3|\Phi^+|$ factors for arbitrary commutative ring of stable rank $1$.

\begin{thm} Let $I$ be an ideal of a commutative ring $R$ such that $\sr(R)\leq n$, where
\begin{itemize}
 \item $n=1$ and $\Phi$ is either classical or of type $\rE_\ell$;
 \item $n=2, 3$ and $\Phi$ is simply laced.
\end{itemize}
Then $\G(\Phi, R, I)$ is a product of at most $q_{n}(\Phi)$ copies of $\SL_{n+1}(R, I)$, where each $q_{n}(\Phi)$, $n=1,2,3$ is a quadratic function of $\rk(\Phi)$ given in Table~XXX.
\end{thm}

@@

\section{Applications}\label{sec:applications}
\subsection{Bounded generation}\label{sec:boundgen}
\input{congrsubgr-boundgen}
\subsection{Subsystem factorizations}\label{sec:subsysfact}
\input{subsystem-factorizations}

>>>>

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative versions of the following three kinds of decompositions:
\begin{itemize}
\item \emph{Gauss decomposition,} i.\,e. a decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

On the other hand, much less is known about the width of relative groups $\G(\Phi, R, I)$. 
It is a classical result of J.~Tits and L.~Vaserstein that the relative elementary group $\E(\Phi, R, I)$ is generated by elements $z_\alpha(s,\xi)=x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, see Theorem~\ref{theorem:Tits-Vaserstein} below.
Denote by $\mathcal{Z}$ the set of all $z_\alpha(s, \xi)$, $s\in I$, $\xi \in R$. 
Our first application of parabolic factorizations is the following estimate of the width of $\G(\Phi, R, I)$ in terms of $\mathcal{Z}$.

@@

Another application of parabolic factorizations developed in this paper are subsystem factorizations in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{NikProdDecomp}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.
In~\cite{VavKovSL2} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, Gauss decomposition (see~\cite{Sm12}) implies a similar factorization for the elementary subgroup $\E(\Phi, R)$ with $3|\Phi^+|$ factors for arbitrary commutative ring of stable rank $1$.

\begin{thm} Let $I$ be an ideal of a commutative ring $R$ such that $\sr(I)\leq N$. Assume that one of the following is true:
\begin{itemize} 
\item $\Phi=\rA_\ell$, $\ell\geq 2$, $\sr(I)\leq 2$;
\item $\Phi=\rC_\ell$, $\ell\geq 2$, $\sr(I)\leq 3$; \end{itemize}
Then $\G(\Phi, R, I)$ equals the product of at most $3|\Phi^+| + A\rk(\Phi) + B$ copies of subgroups of type $\G(\rA_1, R, I)$. \end{thm}

Yet another application of parabolic factorizations concerns subsystem factorizations of Chevalley groups formulated in terms of submaximal subgroups of type $\rA_\ell$.
In case when $R=I$ is a field this question was studied by N.~Nikolov in \cite{NikProdDecomp}. 
\begin{thm*} Let $G$ be a classical Chevalley group (of normal or twisted type) of rank $n$ over a finite field.
Then $G$ equals the product of at most $200$ conjugates of $\SL_n$. \end{thm*}

@@

@@

\section{Applications}\label{sec:applications}
\subsection{Bounded generation}\label{sec:boundgen}
\input{congrsubgr-boundgen}
%\subsection{Subsystem factorizations}\label{sec:subsysfact}
%\input{subsystem-factorizations}
